Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-12 appendix.tex
Still a few points of confusion :pooh:

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
For the binary symmetric model $\alphabet=\{0,1\}$.
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We define an ``ancestral state split'' in a similar way to a site split, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip---\emph{not} internal node---is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
%das: done

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
%das: yah, agreed
Similarly we enumerate ancestral state splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Note
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

@@

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the entropy and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
%EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
%das: I added the table of general generating probabilities, but it's pretty straightforward I think to see below the second derivative of the likelihood will be negative on the whole interval, implying we have a unique maximum.
%I think we're good.
%Let me know if I should say anything about the continuity of the likelihood, but I think it's obvious this gets us to where we need to be?
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, and the map $g(x_1) \mapsto -g(x_1)\log g(x_1)$.
We restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
By straightforward calculus, the second derivative of this map is
\[
-g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.
%EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
%das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?

We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

%EM Shouldn't this read "obtaining an upper bound"?
%das: yes

@@

%EM Shouldn't this be
%&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)

@@

%EM And this be
%+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
%das: I don't think so---the first two lines starting at the equals sign are bringing together the first four lines starting above at the <= sign.
%Should I remove this intermediate equation?
%I thought it would help clarify but maybe not?

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
%EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
%das: is this better? there were a few other typos, so I changed "site split" to "ancestral state split" where needed and updated the constant definitions
For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$, yielding the upper bound
\begin{align}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*).
\end{split}
\label{eq:farris-upper-bound-empty}
\end{align}

@@

We now need to check that
\[
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
\]
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
%EM Using what about t^*?
%das: updated
Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
\[
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
\]
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
\begin{split}
& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
\end{split}
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
For the binary symmetric model, the alphabet $\alphabet$ is $\{0,1\}$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$, commonly called a ``site split.''
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,k}$ be the label of the $k$th tip in the $i$th alignment column.
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
Similarly we enumerate ancestral state splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

@@

%EM There is a latex package for that, which I tried out and then removed for some reason. It's no bigs.

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
%EM Also, I tried to make the connection between the divergence and g a little more explicit.
The second derivative of the divergence term $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t}) = -g(x_1)\log g(x_1)$ is
\[
-g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.

We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
%EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1. Perhaps add a sentence here?
Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

@@

@@

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$, yielding the upper bound
\begin{align}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*).
\end{split}
\label{eq:farris-upper-bound-empty}
\end{align}

@@

We now need to check that
\[
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
\]
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
\[
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
\]
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
\begin{split}
& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
\end{split}
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.
